Revisiting the Nos\'{e}-Hoover chain thermostat in the Rigid Bodies Molecular Dynamics

### 2017-04-04 rigid_bodies.tex
por se me assaltam

Molecular Simulation, being the computational realization of Statistical Mechanics\cite{Tuckerman2010}, involves several types of approximations in order to mimic ``real'' physical systems. Naively, one would expect those approximations to only produce deviation from experimental values. However, they actually may give rise to systematic errors introducing non-physical behaviour. For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics, disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property\cite{Eastwood_2010}. In a previous paper,\citep{Abreu_2017} we reported the same artificiality to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics. That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.

The relevance of rigid-body Molecular Dynamics stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water, but also it is a simple coarse-grain strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules\cite{Knorowski2012, Patra2013}. As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj2005} which is implemented in the softwares packages \textit{blue}-HOOMD and LAMMPS\cite{Plimpton1995}, has been applied to simulate small molecules\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2}, as well as membranes~\cite{Bucior_2012}, molecular motors\cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles\cite{Patra_2014}.

The Nos\'{e}-Hoover chain thermostat,\cite{Martyna1992} routinely used in Molecular Dynamics, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables. The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with what backward error analysis is not feasible anymore. Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi2001,Ezra2004,Ezra2006}

In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body Molecular Dynamics in the canonical ensemble. The starting point is our previous work on the microcanonical ensemble,\citep{Abreu_2017} which employs the unit quaternion representation for the rotational degrees of freedom. Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value. The analysis of the integrators will be rather heuristic, relying on comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane1987} which is exact in the sense that truncation errors do not affect its accuracy. We also include results from simulations using the conventional SHAKE\cite{Ryckaert1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.

The paper is organized as follows. In Sec.~\ref{sec:theory} we briefly review the equations of motion in the microcanonical ensemble, which are the basis for developing the corresponding dynamics in the canonical ensemble. In Sec.~\ref{sec:numerical_solvers} we devise the corresponding numerical integrators, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

\section{Theory}
\label{sec:theory}
Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we have reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble. In the aforementioned paper, we introduced an analytical solution for torque-free rotations, which would be particularly useful if any backward error analysis\cite{Hairer_2006} is to be carried out. However, the  Nos\'{e}-Hoover chain thermostat introduces non-Hamiltonian components, so any backward error analysis it is not feasible anymore. Given that, in this paper, the thermostatted version of the equations of motion will be generated considering the approximate solution for the rotational motion, which is also a mathematical simplification of the formulation introduced by Miller~\textit{et al.}\cite{Miller2002}, and exhibits a precision notably similar to that achieved using the analytical solution. In the next subsection, we will skip over the mathematical details and restrict ourselves to some fundamental expressions on which the dynamics in the canonical (NVT) ensemble will be developed. For full details in the NVE ensemble the reader is referred to our previous paper.\cite{Abreu_2017}

@@

When describing the motion of rigid bodies, first recall that the translational and rotational degrees of freedom are decoupled. Mathematically, the position of a particle $j$ constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the center of mass position of the body, $\vt d_j$ is the particle position in the body-fixed frame of reference, and $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix $\mt A$, which in turn is a function of the unit quaternions $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$. In terms of $\vt q$, the dynamics must satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\dot{\vt q}} = 0$.

The matrix $\mt A$ is computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation*}
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray},
\end{equation*}
which admits the following factorization:
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C},
\end{equation} 
where
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}
Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} a convenient factorization of matrix $\mt A$. This factorization provides a relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components, which is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}
The equation above avoids the introduction of a ficticious angular velocity component, as required in the formulation of Miller~\textit{et al.}\cite{Miller2002} An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation. For the sake of brevity, the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) might be omitted, when such argument is $\vt q$.

In the Hamiltonian approach, we employ the linear momentum $\vt p = m \dot{\vt r}$, conjugated to $\vt r$, as well as the momentum $\vt \pi= 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega$, conjugated to $\vt q$\cite{Goldstein2002}. In this equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a digonal matrix in the principal frame. For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Abreu_2017}

@@

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r}_i = \partial \mathcal{H} / \partial \vt p_i$, $\dot{\vt p}_i = -\partial \mathcal{H} / \partial \vt r_i$, $\dot{\vt q}_i = \partial \mathcal{H} / \partial \vt \pi_i$, and $\dot{\vt \pi}_i = -\partial \mathcal{H} / \partial \vt q_i$.\cite{Goldstein2002}. For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r_i} = -\vt F_i = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference and thus $\vt F_i$ is the resultant force on the body. As demonstrated in Abreu~\textit{et al.},\citep{Abreu_2017} $\partial{U}/\partial \vt q_i$ is given by
\[
\diff{U}{\vt q_i} = -2 \mt C_i \vt \tau_i = -2 \mt C_i \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau_i$ is the resultant torque exerted on the body and ${\vt \delta}_j$ is the position of the particle $j$, both expressed in the space-fixed frame.  The position ${\vt \delta}_j$ is related to ${\vt d}_j$ by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} After carrying out the required differentiations of $\mathcal{H}$  we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r}_i = \frac{1}{m_i} \vt p_i \\
&\dot{\vt p}_i = \mt F_i \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:EDO_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i \label{eq:EDO_pi}
\end{align}
\end{subequations}
with $\mt \Omega_i = \mt B_i(\vt \pi_i)$.

The approximate solution for a complete rotation is obtained by opening the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ employing each entry of $\vt \omega$: 
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the following permutation matrices:
\[
\hat{\mt B}_1 = \begin{smallarray}{cccc}
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
 0 &  0 &  0 &  1 \\
 0 &  0 & \text{-}1 &  0 \\
\end{smallarray} \quad
\hat{\mt B}_2 = \begin{smallarray}{cccc}
 0 &  0 & \text{-}1 &  0 \\
 0 &  0 &  0 & \text{-}1 \\
 1 &  0 &  0 &  0 \\
 0 &  1 &  0 &  0 \\
\end{smallarray} \quad
\hat{\mt B}_3 = \begin{smallarray}{cccc}
 0 &  0 &  0 & \text{-}1 \\
 0 &  0 &  1 &  0 \\
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
\end{smallarray}
\]
The new form of the equations of motion in Eqs.~\eqref{eq:EDO_q} and \eqref{eq:EDO_pi} are then:
\begin{subequations}
\label{eq:EDO_system_alternative}
\begin{align}
&\dot{\vt q}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt q_i \label{eq:edo_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt \pi_i + 2 \mt C_i \vt \tau_i \label{eq:edo_pi}
\end{align}
\end{subequations}

In this context, the Hamiltonian of the rigid body can be rewritten as
\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} =  U(\vt r, \vt q) +  \sum_{i=1}^N \left( \frac{\tr{\vt p_i} \vt p_i}{2m_i} + \frac{1}{2} \sum_{k=1}^3 I_{ik} \omega_{ik}^2\right),
\end{equation}
from which we recognize partial Hamiltonians, each of one generates equations of motion with known analytical solutions. In particular, we have $\mathcal{H}_b = U(\vt r, \vt q)$, $\mathcal{H}_{t} = \frac{1}{2m}\tr{\vt p} \vt p$, and $\mathcal{H}_k = (1/2) I_k \omega_k^2$. Of coarse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$. The NVE dynamics of this system is governed by a Liouville operator given by\cite{Abreu_2017}
\begin{equation}
\label{eq:full_operator}
i\!L_{NVE} = i\!L_t + i\!L_b + i\!L_r,
\end{equation}
where the individual operators are defined as follows:
\[
\begin{split}
&boost:\\
&i\!L_b = \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
&translation: \\
& i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
&uniaxial \; rotation:\\
& i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right),
\end{split}
\]
using $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete torque-free rotation. As explained in our previous paper,\cite{Abreu2017} those individual operators can be applied analytically according to
\[
\begin{split}
&boost:\\
&{\vt p} = {\vt p}_0 + {\vt F}_0 t \\
&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t\\
&translation:\\
&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0\\
&uniaxial \; rotation:\\
&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
\end{split}
\]

Applying the Trotter-susz The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b},
\end{equation}
Update of forces and torques is required only after the free translation and free rotation have been carried out. Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is splitted as
\begin{align*}
\label{eq:splitting_rot}
e^{\Delta t i\!L_r} =  e^{\frac{\Delta t}{2} i\!L_3} e^{\frac{\Delta t}{2} i\!L_2} e^{\Delta t i\!L_1} e^{\frac{\Delta t}{2} i\!L_2} e^{\frac{\Delta t}{2} i\!L_3},
\end{align*}
where ${\Delta t}$ is the time step size.

equipartion of energy

@@

%\begin{equation}
%\label{eq:probability_ratio_kin}
%\frac{P_{kin}(E/\beta_2)}{P_{kin}(E/\beta_1)} = \frac{Q_{kin}(\beta_1)}{Q_{kin}(\beta_2)}  \exp\left[(-%\beta_2 - \beta_1)E_{kin}\right]
%\end{equation}
\end{document}
\grid

>>>>

Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman2010} involves several types of approximations in order to mimic ``real'' physical systems. Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field. However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour. For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Abreu_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics. That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.

The relevance of rigid-body Molecular Dynamics stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski2012, Patra2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton1995}, has been applied to simulate small molecules,\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}

The Nos\'{e}-Hoover chain thermostat,\cite{Martyna1992} routinely used in Molecular Dynamics, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables. The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is not longer feasible. Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi2001,Ezra2004,Ezra2006}

In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body Molecular Dynamics in the canonical (NVT) ensemble. The starting point is our previous work on the microcanonical ensemble,\citep{Abreu_2017} which employs the unit quaternion representation for the rotational degrees of freedom. Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value. The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane1987} which is exact in the sense that truncation errors do not affect its accuracy. We also include results from simulations using the conventional SHAKE\cite{Ryckaert1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.

The paper is organized as follows. In Sec.~\ref{sec:theory} we briefly review the equations of motion in the microcanonical (NVT) ensemble, which are the basis for developing the corresponding dynamics in the NVT ensemble. In Sec.~\ref{sec:numerical_solvers} we devise the corresponding numerical integrators, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

\section{Theory}
\label{sec:theory}

Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we introduced symplectic numerical integrators for MD in the NVE ensemble, for which we derived an analytical, as well as an approximate solution for torque-free rotations. This last approach is a reformulation of the method introduced by Miller~\textit{et al.},\cite{Miller2002} and given that it demonstrated to be as accurate as the exact solution, the thermostatted version of the equations of motion will be derived considering the approximate solution. In the next subsection, we will skip over the mathematical derivations and restrict ourselves to some fundamental expressions.

@@

When describing the motion of rigid bodies, first recall that the translational and rotational degrees of freedom are decoupled. Mathematically, the position of a particle $j$ constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the center of mass position of the body, $\vt d_j$ is the particle position in the body-fixed frame of reference, and $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix $\mt A$, which in turn is a function of the unit quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$. In terms of $\vt q$, the dynamics must satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\dot{\vt q}} = 0$.

The matrix $\mt A$ is computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation*}
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray},
\end{equation*}
which admits the following factorization:
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C},
\end{equation} 
where
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}
Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} a convenient factorization of matrix $\mt A$ and most important, it provides the mathematical relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components, which is given by
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}
 For the sake of brevity, the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) might be omitted, when such argument is $\vt q$.

In the Hamiltonian approach, we employ the linear momentum $\vt p = m \dot{\vt r}$, conjugated to $\vt r$, as well as the momentum $\vt \pi= 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega$, conjugated to $\vt q$\cite{Goldstein2002}. In these equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a digonal matrix in the principal frame. For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Abreu_2017}

@@

The equations of motion are derived from the Hamiltonian Eq.\eqref{eq:H_NVE} by $\dot{\vt r}_i = \partial \mathcal{H} / \partial \vt p_i$, $\dot{\vt p}_i = -\partial \mathcal{H} / \partial \vt r_i$, $\dot{\vt q}_i = \partial \mathcal{H} / \partial \vt \pi_i$, and $\dot{\vt \pi}_i = -\partial \mathcal{H} / \partial \vt q_i$.\cite{Goldstein2002}. For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r_i} = -\vt F_i = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference, and thus $\vt F_i$ is the resultant force on the body. As demonstrated in Abreu~\textit{et al.},\citep{Abreu_2017} $\partial{U}/\partial \vt q_i$ is given by
\[
\diff{U}{\vt q_i} = -2 \mt C_i \vt \tau_i = -2 \mt C_i \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau_i$ is the resultant torque exerted on the body, and ${\vt \delta}_j$ $=$ $\tr{\mt A} {\vt d}_j$ is the position of the particle $j$, both expressed in the space-fixed frame. After carrying out the required differentiations of Eq.\eqref{eq:H_NVE}, we obtain the following system of equations for the rigid body motion:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r}_i = \frac{1}{m_i} \vt p_i \\
&\dot{\vt p}_i = \mt F_i \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:EDO_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i \label{eq:EDO_pi}
\end{align}
\end{subequations}
with $\mt \Omega_i = \mt B_i(\vt \pi_i)$.

The approximate solution for rotation is derived by opening the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ from each entry of $\vt \omega$: 
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the following permutation matrices:
\[
\hat{\mt B}_1 = \begin{smallarray}{cccc}
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
 0 &  0 &  0 &  1 \\
 0 &  0 & \text{-}1 &  0 \\
\end{smallarray} \quad
\hat{\mt B}_2 = \begin{smallarray}{cccc}
 0 &  0 & \text{-}1 &  0 \\
 0 &  0 &  0 & \text{-}1 \\
 1 &  0 &  0 &  0 \\
 0 &  1 &  0 &  0 \\
\end{smallarray} \quad
\hat{\mt B}_3 = \begin{smallarray}{cccc}
 0 &  0 &  0 & \text{-}1 \\
 0 &  0 &  1 &  0 \\
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
\end{smallarray}
\]
The new form of the equations of motion in Eqs.~\eqref{eq:EDO_q} and \eqref{eq:EDO_pi} are then:
\begin{subequations}
\label{eq:EDO_system_alternative}
\begin{align}
&\dot{\vt q}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt q_i \label{eq:edo_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt \pi_i + 2 \mt C_i \vt \tau_i \label{eq:edo_pi}
\end{align}
\end{subequations}

In this context, the Hamiltonian of Eq.\eqref{eq:H_NVE} can be rewritten as
\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} =  U(\vt r, \vt q) +  \sum_{i=1}^N \left( \frac{\tr{\vt p_i} \vt p_i}{2m_i} + \frac{1}{2} \sum_{k=1}^3 I_{ik} \omega_{ik}^2\right),
\end{equation}
from which we recognize partial Hamiltonians, each of one generates equations of motion with known analytical solutions. In particular, we have $\mathcal{H}_b = U(\vt r, \vt q)$, $\mathcal{H}_{t} = \frac{1}{2m}\tr{\vt p} \vt p$, and $\mathcal{H}_k = (1/2) I_k \omega_k^2$. Of corse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$. The NVE dynamics of this system is governed by a Liouville operator given by\cite{Abreu_2017}
\begin{equation}
\label{eq:full_operator}
i\!L_\text{NVE} = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k,
\end{equation}
where the individual operators are defined as follows:
\[
\begin{split}
&boost:\\
&i\!L_b = \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right) \\
&translation: \\
& i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i} \\
&uniaxial \; rotation:\\
& i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right)
\end{split}
\]
The individual operators can be evaluated analytically according to
\[
\begin{split}
&boost:\\
&{\vt p} = {\vt p}_0 + {\vt F}_0 t \\
&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t\\
&translation:\\
&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0\\
&uniaxial \; rotation:\\
&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
\end{split}
\]
Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is factorized as
\begin{align*}
\label{eq:splitting_rot}
e^{\Delta t i\!L_r} =  e^{\frac{\Delta t}{2} i\!L_3} e^{\frac{\Delta t}{2} i\!L_2} e^{\Delta t i\!L_1} e^{\frac{\Delta t}{2} i\!L_2} e^{\frac{\Delta t}{2} i\!L_3},
\end{align*}
where ${\Delta t}$ is the time step size. The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L_\text{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
\end{equation}

@@

%\begin{equation}
%\label{eq:probability_ratio_kin}
%\frac{P_{kin}(E/\beta_2)}{P_{kin}(E/\beta_1)} = \frac{Q_{kin}(\beta_1)}{Q_{kin}(\beta_2)}  \exp\left[(-%\beta_2 - \beta_1)E_{kin}\right]
%\end{equation}

%The equation above avoids the introduction of a ficticious angular velocity component, as required in the formulation of Miller~\textit{et al.}\cite{Miller2002} An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
\end{document}
\grid
